---
title: "Eviction in California: A Collaborative Data Analysis"
subtitle: "DataHack 2026 — In-Class Exercise"
date: "Spring 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

---

# Overview

In this exercise you will work in **pairs** to analyze eviction filing data across
California counties (2000–2018), using data from the
[Eviction Lab](https://evictionlab.org/) at Princeton University.

By the end of class you will have produced:

- A **choropleth map** of eviction rates across all 58 California counties
- A **time series** comparing coastal vs. inland counties over time
- **Summary tables** and descriptive statistics by group
- A simple **OLS regression** estimating differences in eviction rates

You will do this as a real collaboration: two people, one GitHub repository,
using branches, commits, pushes, and pulls.

---

# GitHub Setup

## Create your pair repository

One student per pair does the following:

1. Go to the course template repository on GitHub
2. Click **"Use this template" → "Create a new repository"**
3. Name it something like `eviction-yournames` and set it to **Public**
4. Once created, go to **Settings → Collaborators** and add your partner

The other student:

1. Accepts the collaborator invitation (check your email)
2. Clones the repository in RStudio:
   **File → New Project → Version Control → Git**, paste the repo URL

Both students open the project in RStudio and verify:

```bash
git status
# Should say: On branch main
```

Set your identity if you have not already:

```bash
git config --global user.name  "Your Name"
git config --global user.email "your@email.com"
```

## Create your branch

**Do this before touching any files.**

**Student A — in the RStudio Terminal:**
```bash
git checkout -b student-a
git push -u origin student-a
```

**Student B — in the RStudio Terminal:**
```bash
git checkout -b student-b
git push -u origin student-b
```

Check you are on the right branch:
```bash
git branch
# The branch with * is your current branch
```

---

# The Data

You are working with two files inside `00_raw_data/`:

**File 1 — `county_court-issued_2000_2018.csv`**
Court-issued eviction filings by county and year. Covers all 58 California counties.
This is Student A's input file.

```
state | county | fips_state | fips_county | year | renting_hh | filings_observed | ...
```

**File 2 — `county_proprietary_valid_2000_2018.csv`**
Proprietary estimates with pre-computed filing rates, threatened rates, and judgement
rates. Covers 16 smaller California counties. This is Student B's input file.

```
cofips | county | state | year | type | filings | filing_rate | threatened_rate | judgement_rate
```

**Key variable you will compute (Student A):**
$$\text{filing rate} = \frac{\text{filings observed}}{\text{renting households}} \times 100$$

This is the percentage of renting households that received an eviction filing
in a given county and year.

---

# Part 1 — Data Cleaning (parallel, ~25 min)

Work independently on your own branch. Open your script and run it
**section by section** — do not source the whole file at once.

---

## Student A: County court-issued data

Open `01_cleaning_scripts/student_A_county_cleaning.R`.

**Step 1 — Load packages and data:**
```{r}
rm(list = ls())
library(pacman)
p_load(tidyverse, tidylog, janitor)

county_raw <- read_csv("00_raw_data/county_court-issued_2000_2018.csv")
glimpse(county_raw)
```

**Step 2 — Filter to California:**
```{r}
ca_county <- county_raw %>%
  filter(state == "California")

n_distinct(ca_county$county)  # should be 58
```

**Step 3 — Clean county names and fix FIPS:**
```{r}
ca_county <- ca_county %>%
  mutate(
    county = str_remove(county, " County"),
    geoid  = str_pad(as.character(fips_county), width = 5, pad = "0")
  )

# Verify: all geoids should start with "06"
unique(ca_county$geoid) %>% head()
```

**Step 4 — Compute eviction filing rate:**
```{r}
ca_county <- ca_county %>%
  mutate(filing_rate = filings_observed / renting_hh * 100)

summary(ca_county$filing_rate)
# Values between 0 and 15 are expected (percentage of renting HH)
```

**Step 5 — Create group indicators:**
```{r}
coastal_counties <- c(
  "Del Norte", "Humboldt", "Mendocino", "Sonoma", "Marin",
  "San Francisco", "San Mateo", "Santa Cruz", "Monterey",
  "San Luis Obispo", "Santa Barbara", "Ventura",
  "Los Angeles", "Orange", "San Diego"
)

ca_county <- ca_county %>%
  mutate(
    coastal  = if_else(county %in% coastal_counties, 1L, 0L),
    post2008 = if_else(year >= 2008, 1, 0)
  )

ca_county %>% count(coastal)
```

**Step 6 — Save:**
```{r}
save(ca_county, file = "02_clean_data/ca_county_evictions.RData")
```

✅ **Commit and push:**
```bash
git add 02_clean_data/ca_county_evictions.RData
git commit -m "Student A: clean county court-issued data for California"
git push origin student-a
```

---

## Student B: County proprietary data

Open `01_cleaning_scripts/student_B_county_cleaning.R`.

**Step 1 — Load packages and data:**
```{r}
rm(list = ls())
library(pacman)
p_load(tidyverse, tidylog, janitor)

prop_raw <- read_csv("00_raw_data/county_proprietary_valid_2000_2018.csv")
glimpse(prop_raw)
```

**Step 2 — Filter to California, keep only observed rows:**
```{r}
ca_prop <- prop_raw %>%
  filter(state == "California") %>%
  filter(type == "observed")

nrow(ca_prop)           # 123 rows
n_distinct(ca_prop$county)  # 16 counties
```

> **Why only 16 counties?** The Eviction Lab's proprietary data has valid
> court records for only 16 small rural California counties. The remaining
> 42 counties (including LA, SF, etc.) will come from Student A's file.

**Step 3 — Clean names and FIPS:**
```{r}
ca_prop <- ca_prop %>%
  mutate(
    county = str_remove(county, " County"),
    geoid  = str_pad(as.character(cofips), width = 5, pad = "0")
  )
```

**Step 4 — Select and rename columns:**
```{r}
# Rename filing_rate → filing_rate_prop to avoid clash when merging with Student A
ca_prop_clean <- ca_prop %>%
  select(
    geoid,
    county,
    year,
    filing_rate_prop = filing_rate,
    threatened_rate,
    judgement_rate
  )

glimpse(ca_prop_clean)
```

**Step 5 — Save:**
```{r}
save(ca_prop_clean, file = "02_clean_data/ca_prop_clean.RData")
```

✅ **Commit and push:**
```bash
git add 02_clean_data/ca_prop_clean.RData
git commit -m "Student B: clean proprietary county data for California"
git push origin student-b
```

---

## Sync: merge both branches into main

Both cleaned files need to be on `main` before the merge script can run.
**Do this together.**

**Student A goes first:**
```bash
git checkout main
git merge student-a
git push origin main
```

**Student B — pull first, then merge:**
```bash
git checkout main
git pull origin main
git merge student-b
git push origin main
```

Both students pull the updated main:
```bash
git pull origin main
```

Verify both files exist:
```bash
ls 02_clean_data/
# ca_county_evictions.RData
# ca_prop_clean.RData
```

---

# Part 2 — Merge (~10 min)

**One student** runs `03_analysis_scripts/00_merge.R`.

This script loads both cleaned files and joins them:

```{r}
load("02_clean_data/ca_county_evictions.RData")  # ca_county (477 rows)
load("02_clean_data/ca_prop_clean.RData")         # ca_prop_clean (123 rows)

ca_county_merged <- ca_county %>%
  left_join(
    ca_prop_clean %>% select(-county),
    by = c("geoid", "year")
  )

# How many rows matched?
ca_county_merged %>%
  summarise(
    total_rows    = n(),
    has_prop_rate = sum(!is.na(filing_rate_prop)),
    has_judgement = sum(!is.na(judgement_rate))
  )
```

> The left join keeps all 477 rows from Student A. Only ~55 rows will match
> with Student B's file (the 16 rural counties). That is expected — the main
> analysis variables (`filing_rate`, `coastal`, `post2008`) are complete
> for all 58 counties.

```{r}
save(ca_county_merged, file = "02_clean_data/ca_county_merged.RData")
```

✅ **Commit and push:**
```bash
git add 02_clean_data/ca_county_merged.RData
git commit -m "merge Student A and B outputs into ca_county_merged"
git push origin main
```

The other student pulls:
```bash
git pull origin main
```

---

# Part 3 — Analysis (parallel, ~30 min)

Switch back to your branch and bring in the merged data:

**Student A:**
```bash
git checkout student-a
git merge main
```

**Student B:**
```bash
git checkout student-b
git merge main
```

---

## Student A: Maps and graphs

Open `03_analysis_scripts/student_A_maps_graphs.R`.

**Step 1 — Load data and get California county geometries:**
```{r}
library(pacman)
p_load(tidyverse, tidylog, janitor, tigris, sf)
options(tigris_use_cache = TRUE)

load("02_clean_data/ca_county_merged.RData")

# Download Census county boundaries (takes ~10 sec first time, then cached)
ca_sf <- counties(state = "CA", cb = TRUE, year = 2018, progress_bar = FALSE)

# GEOID in ca_sf matches our geoid column ("06001" format)
head(ca_sf$GEOID)
```

**Step 2 — Compute county averages and join geometry:**
```{r}
ca_avg <- ca_county_merged %>%
  group_by(geoid, county, coastal) %>%
  summarise(avg_filing_rate = mean(filing_rate, na.rm = TRUE), .groups = "drop")

ca_map <- ca_sf %>%
  left_join(ca_avg, by = c("GEOID" = "geoid"))
```

**Step 3 — Map 1: average filing rate across all years:**
```{r}
ggplot(ca_map) +
  geom_sf(aes(fill = avg_filing_rate), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(name = "Avg filing rate\n(% renting HH)",
                       option = "magma", direction = -1) +
  labs(title = "Average Eviction Filing Rate by County",
       subtitle = "California, 2010–2017",
       caption = "Source: Eviction Lab") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave("04_graphs/map_avg_filing_rate.png", width = 7, height = 8, dpi = 150)
```

**Step 4 — Map 2: filing rate in 2010:**
```{r}
ca_map_2010 <- ca_sf %>%
  left_join(
    ca_county_merged %>% filter(year == 2010) %>% select(geoid, filing_rate),
    by = c("GEOID" = "geoid")
  )

ggplot(ca_map_2010) +
  geom_sf(aes(fill = filing_rate), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(name = "Filing rate\n(% renting HH)",
                       option = "magma", direction = -1) +
  labs(title = "Eviction Filing Rate by County, 2010",
       subtitle = "California — one year after the housing crisis",
       caption = "Source: Eviction Lab") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave("04_graphs/map_filing_rate_2010.png", width = 7, height = 8, dpi = 150)
```

**Step 5 — Time series: coastal vs. inland:**
```{r}
ts_data <- ca_county_merged %>%
  mutate(group = if_else(coastal == 1, "Coastal", "Inland")) %>%
  group_by(year, group) %>%
  summarise(mean_rate = mean(filing_rate, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  filter(n_distinct(group) == 2) %>%   # only years where BOTH groups have data
  ungroup()

ggplot(ts_data, aes(x = year, y = mean_rate, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 2008, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Coastal" = "#2166ac", "Inland" = "#d6604d")) +
  labs(title = "Eviction Filing Rates: Coastal vs. Inland Counties",
       subtitle = "California, 2010–2017",
       x = NULL, y = "Mean filing rate (% renting HH)", color = NULL,
       caption = "Source: Eviction Lab") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave("04_graphs/timeseries_coastal_inland.png", width = 8, height = 5, dpi = 150)
```

**Step 6 — Bar chart: top 10 counties:**
```{r}
top10 <- ca_avg %>%
  slice_max(avg_filing_rate, n = 10) %>%
  mutate(county = fct_reorder(county, avg_filing_rate),
         group  = if_else(coastal == 1, "Coastal", "Inland"))

ggplot(top10, aes(x = avg_filing_rate, y = county, fill = group)) +
  geom_col() +
  scale_fill_manual(values = c("Coastal" = "#2166ac", "Inland" = "#d6604d")) +
  labs(title = "Top 10 Counties by Average Eviction Filing Rate",
       subtitle = "California, 2010–2017",
       x = "Average filing rate (% renting HH)", y = NULL, fill = NULL,
       caption = "Source: Eviction Lab") +
  theme_minimal() +
  theme(plot.background = element_rect(fill = "white", color = NA))

ggsave("04_graphs/top10_counties.png", width = 7, height = 5, dpi = 150)
```

✅ **Commit and push:**
```bash
git add 04_graphs/
git commit -m "Student A: maps and graphs for CA eviction data"
git push origin student-a
```

---

## Student B: Tables and regression

Open `03_analysis_scripts/student_B_tables_regression.R`.

**Step 1 — Load data:**
```{r}
library(pacman)
p_load(tidyverse, tidylog, janitor, modelsummary)

load("02_clean_data/ca_county_merged.RData")
```

**Step 2 — Summary statistics table:**
```{r}
summary_stats <- ca_county_merged %>%
  select(filing_rate, filings_observed, renting_hh, coastal, post2008) %>%
  summarise(across(everything(),
    list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE),
         min  = ~min(.x,  na.rm = TRUE), max = ~max(.x, na.rm = TRUE)),
    .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(), names_to = c("variable", "stat"), names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value)

print(summary_stats)
write_csv(summary_stats, "05_tables/table_summary_stats.csv")
```

**Step 3 — Mean filing rate by group:**
```{r}
means_by_group <- ca_county_merged %>%
  mutate(group  = if_else(coastal  == 1, "Coastal",   "Inland"),
         period = if_else(post2008 == 1, "Post-2008", "Pre-2008")) %>%
  group_by(group, period) %>%
  summarise(mean_rate = mean(filing_rate, na.rm = TRUE),
            sd_rate   = sd(filing_rate,   na.rm = TRUE),
            n_obs     = n(), .groups = "drop")

print(means_by_group)
write_csv(means_by_group, "05_tables/table_means_by_group.csv")
```

**Step 4 — T-test: are coastal and inland rates significantly different?**
```{r}
t.test(
  ca_county_merged %>% filter(coastal == 1) %>% pull(filing_rate),
  ca_county_merged %>% filter(coastal == 0) %>% pull(filing_rate)
)
```

**Step 5 — OLS regression (three models):**
```{r}
m1 <- lm(filing_rate ~ coastal,            data = ca_county_merged)
m2 <- lm(filing_rate ~ coastal + post2008, data = ca_county_merged)
m3 <- lm(filing_rate ~ coastal * post2008, data = ca_county_merged)

# Model 1: raw coastal gap
# Model 2: controlling for the time period
# Model 3: interaction — did the 2008 crisis affect coastal and inland differently?

modelsummary(
  list("Model 1" = m1, "Model 2" = m2, "Model 3" = m3),
  stars       = TRUE,
  gof_map     = c("nobs", "r.squared", "adj.r.squared"),
  coef_rename = c("coastal"          = "Coastal county",
                  "post2008"         = "Post-2008",
                  "coastal:post2008" = "Coastal × Post-2008"),
  title = "OLS: Eviction Filing Rate",
  notes = "Source: Eviction Lab. Unit of observation: county-year."
)
```

**Step 6 — Save regression output:**
```{r}
sink("05_tables/table_regression.txt")
modelsummary(list("Model 1" = m1, "Model 2" = m2, "Model 3" = m3), stars = TRUE)
sink()
```

✅ **Commit and push:**
```bash
git add 05_tables/
git commit -m "Student B: summary tables and regression"
git push origin student-b
```

---

# Part 4 — Final merge (~5 min)

Bring everything together on `main`.

**Student A:**
```bash
git checkout main
git merge student-a
git push origin main
```

**Student B — pull first:**
```bash
git checkout main
git pull origin main
git merge student-b
git push origin main
```

Check the full commit history:
```bash
git log --oneline
```

You should see commits from both students across cleaning, merging, and analysis.

---

# Checklist

- [ ] `02_clean_data/ca_county_evictions.RData` — Student A cleaning ✓
- [ ] `02_clean_data/ca_prop_clean.RData` — Student B cleaning ✓
- [ ] `02_clean_data/ca_county_merged.RData` — merge step ✓
- [ ] `04_graphs/` — 4 `.png` files from Student A ✓
- [ ] `05_tables/` — 3 output files from Student B ✓
- [ ] Both students have commits visible in `git log --oneline` ✓
- [ ] `main` is up to date with all work merged in ✓

---

*Data source: Eviction Lab, Princeton University. [evictionlab.org](https://evictionlab.org/)*
